Interfacing logic programming and Coq

### 2020-10-23 papers/types2020/pbt.tex
some metanotes on how what to write in sec 3 and 4

>>>>

\begin{metanote}
  dump from previous paper
\end{metanote}

When using a focused proof system for logic extended with fixed
points, such as  employed in Bedwyr~\cite{baelde07cade} and described
in~\cite{baelde12tocl,heath17linearity}, proofs of formulas such as
\[
  \exists x [(\tau(x)\wedge P(x)) \wedge \neg Q(x)]
  \tag{*}\label{eq:full}
\]
are a single \emph{bipole}: that is, when reading a proof bottom up, a
positive phase is followed on all its premises by a single negative
phase that completes the proof.%
\footnote{The reader familiar with focusing will understand that there
are two ``polarized'' conjunctions, written in linear logic as
$\otimes$ and $\with$ or in classical and intuitionistic logics as
$\wedgep$ and $\wedgen$, respectively.  In this paper, we use simply
$\wedge$ to denote the positive biased conjunction.}
%
In particular, the positive phase corresponds to the \emph{generation} phase
and the negative phase corresponds to the \emph{testing} phase.
%

Instead of giving a full focused proof system of a logic including
fixed points (since, as we will argue, that proof system will not, in
fact, be needed to account for PBT ), we offer the following analogy.
%
Suppose that we are given a finite search tree and we are asked to
prove that there is a secret located in one of the nodes of that
tree.
%
A proof that we have found that secret can be taken to be a
description of the path to that node from the root of the tree: that
path can be seen as the proof certificate for that claim.
%
On the other hand, a proof that no node contains the secret is a
rather different thing: here, one expects to use a blind and
exhaustive search (via, say, depth-first or breath-first search) and
that the result of that search never discovers the secret.
%
A proof of this fact requires \emph{no} external information: instead it
requires a lot of computation involved with exploring the tree until
exhaustion.
%
This follows the familiar pattern where the positive (generate) phase
requires external information while the negative (testing) phase
requires none.
%
Thus, a proof certificate for $(\ref{eq:full})$ is also a proof
certificate for 
\[
  \exists x [\tau(x)\wedge P(x)].
  \tag{**}\label{eq:short}
\]
Such a certificate would contain the witness (closed) term $t$ for the
existential quantifier and sufficient information to confirm that
$P(t)$ can be proved (a proof of a typing judgment such as $\tau(t)$
is usually trivial).
%
%In short\ednote{AM: a bit repetitious}, a proof certificate for
%$(\ref{eq:short})$ can play the part 
%of a proof certificate also for $(\ref{eq:full})$, if we also use
%negation-as-finite-failure  with $Q(t)$.
%  ****
Since a proof certificate for the existence-of-a-counterexample formula
$(\ref{eq:full})$ can be taken as a proof certificate of
$(\ref{eq:short})$, then we only need to consider proof certificates
for Horn clause programs.
%
%Since the logic of Horn clauses is rather simple,
We illustrate next what such proofs and proof certificates look like
for the rather simple logic of Horn clauses.

### 2020-10-23 papers/types2020/sec3.tex
some metanotes on how what to write in sec 3 and 4

\subsection{Focused proof systems}
\label{ssec:focused}

@@

When using a focused proof system for logic extended with fixed
points, such as  employed in Bedwyr~\cite{baelde07cade} and described
in~\cite{baelde12tocl,heath17linearity}, proofs of formulas such as
\[
  \exists x [(\tau(x)\wedge P(x)) \wedge \neg Q(x)]
  \tag{*}\label{eq:full}
\]
are a single \emph{bipole}: that is, when reading a proof bottom up, a
positive phase is followed on all its premises by a single negative
phase that completes the proof.%
\footnote{The reader familiar with focusing will understand that there
are two ``polarized'' conjunctions, written in linear logic as
$\otimes$ and $\with$ or in classical and intuitionistic logics as
$\wedgep$ and $\wedgen$, respectively.  In this paper, we use simply
$\wedge$ to denote the positive biased conjunction.}
%
In particular, the positive phase corresponds to the \emph{generation} phase
and the negative phase corresponds to the \emph{testing} phase.
%

Instead of giving a full focused proof system of a logic including
fixed points (since, as we will argue, that proof system will not, in
fact, be needed to account for PBT ), we offer the following analogy.
%
Suppose that we are given a finite search tree and we are asked to
prove that there is a secret located in one of the nodes of that
tree.
%
A proof that we have found that secret can be taken to be a
description of the path to that node from the root of the tree: that
path can be seen as the proof certificate for that claim.
%
On the other hand, a proof that no node contains the secret is a
rather different thing: here, one expects to use a blind and
exhaustive search (via, say, depth-first or breath-first search) and
that the result of that search never discovers the secret.
%
A proof of this fact requires \emph{no} external information: instead it
requires a lot of computation involved with exploring the tree until
exhaustion.
%
This follows the familiar pattern where the positive (generate) phase
requires external information while the negative (testing) phase
requires none.
%
Thus, a proof certificate for $(\ref{eq:full})$ is also a proof
certificate for 
\[
  \exists x [\tau(x)\wedge P(x)].
  \tag{**}\label{eq:short}
\]
Such a certificate would contain the witness (closed) term $t$ for the
existential quantifier and sufficient information to confirm that
$P(t)$ can be proved (a proof of a typing judgment such as $\tau(t)$
is usually trivial).
%
%In short\ednote{AM: a bit repetitious}, a proof certificate for
%$(\ref{eq:short})$ can play the part 
%of a proof certificate also for $(\ref{eq:full})$, if we also use
%negation-as-finite-failure  with $Q(t)$.
%  ****
Since a proof certificate for the existence-of-a-counterexample formula
$(\ref{eq:full})$ can be taken as a proof certificate of
$(\ref{eq:short})$, then we only need to consider proof certificates
for Horn clause programs.
%
%Since the logic of Horn clauses is rather simple,
We illustrate next what such proofs and proof certificates look like
for the rather simple logic of Horn clauses.

@@

%% \begin{metanote}
%%   AM: I know this is well known and boring, but using \texttt{prog A
%%     G} in Fig \ref{fig:augmented} is imprecise --- in the comment of
%%   next figure we mention \texttt{A :- G} and we gloss over the
%%   instantiation of \texttt{prog A' G} to match the goal \texttt{A},
%%   which we assume to be ground, given the exists rule
%% \end{metanote}

%% DM I've modified the text a bit to hopefully improve the state of
%% things here.  The formal proof theory is all about closed
%% formulas.  The prolog implementation allows open formulas.  This
%% difference invades most aspects of proof checking, not just the
%% prog predicate.

% DM We should find a place to smooth over this issue (which is not
% hard, I think).

\begin{figure}
\newcommand{\XXi}{{\color{blue}{\Xi}}}
\[
\infer{\XXi\vdash t = t}
      {\eqExpert{\XXi}}
\qquad
\infer{\XXi\vdash \true}
      {\trueExpert{\XXi}}
\qquad
\infer{\XXi\vdash G_1\vee G_2}
      {\XXi'\vdash G_i\qquad \orExpert{\XXi}{\XXi'}{i}}
\qquad
\infer{\XXi\vdash \exists x. G}
      {\XXi'\vdash G[t/x]\qquad \someExpert{\XXi}{\XXi'}{t}}
\]
\vskip -6pt
\[
\infer{\XXi\vdash G_1\wedge G_2}
      {\XXi_1\vdash G_1\quad \XXi_2\vdash G_2\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}}
\qquad
\infer{\XXi\vdash A}
      {\XXi'\vdash G \quad (A~\hbox{\tt :-}~G)\in\instan\Pscr
                     \quad \unfoldExpert{\XXi}{\XXi'}}
\]
\caption{A proof system augmented with proof certificates and
  additional ``expert'' premises.}
\label{fig:augmented}
TODO
%\lstinputlisting[linerange={sigs-end}]{code/kernel.sig}
%\lstinputlisting[linerange={check-end}]{code/kernel.mod}
\caption{A simple proof checking kernel.}
\label{fig:kernel}
\end{figure}

Figure~\ref{fig:augmented} contains a simple proof system for Horn
clause provability in which each inference rule is augmented with an
additional premise involving an \emph{expert predicate}, a
certificate $\Xi$, and possibly continuations of certificates ($\Xi'$,
$\Xi_1$, $\Xi_2$) with extracted information from certificates (in the
case of $\vee$ and $\exists$).
%
%
The premise $(A~\hbox{\tt :-}~G)\in\instan{\Pscr}$ in the last
inference rule of Figure~\ref{fig:augmented} states that the logic
programming clause $(A~\hbox{\tt :-}~G)$ is a ground instance of some
clause in a fixed program $\Pscr$.
%
Although this proof system is a focused proof system, the
richness of focusing is not apparent in this simplified 
setting: thus, we drop the adjective ``focused'' from this point forward.

Figure~\ref{fig:kernel} contains the \lP implementation of the
inference rules in Figure~\ref{fig:augmented}: here the infix
turnstile $\vdash$ symbol is replaced by the \lst<i{check} predicate.
%
Notice that it is easy to show that no matter how the expert predicates
are defined, if the goal \lsti{check Cert B} is provable in \lP then
\lsti{B} must be a sound consequence of the program clauses stored in
the \lsti{prog} predicate (which provides a natural implementation of
the premise $(A~\hbox{\tt :-}~G)\in\instan{\Pscr}$.
% AM: not there
% in Figure~\ref{fig:kernel}).

@@

\begin{figure}
\lstinputlisting[linerange={resources-end}]{code/fpcs.sig}
\lstinputlisting[linerange={resources-end}]{code/fpcs.mod}
\caption{Two FPCs that describe proofs that are limited in either
  height or in size.}
\label{fig:resources}
\end{figure}

@@

Various  additional examples and experiments using the pairing of FPCs
can be found in~\cite{Pair}. Using similar techniques, it is possible to define FPCs that target
specific types for special treatment: for example, when generating
integers, only (user-defined) small integers could be inserted into
counterexamples.

>>>>

\begin{metanote}
  (AM): I would start with the logical reconstruction of \texttt{check}, adding the LJT-style proof terms to fig \ref{fig:augmented} and then reverse engineer what we want/need to say about focusing and FPC. We \textbf{do} not talk about PBT yet, just logic programming (with proof terms). In sec 4 we package \texttt{check} in a PBT query and explain how \texttt{check} works for term generation.
\end{metanote}
\subsection{Focused proof systems}
\label{ssec:focused}

@@

@@

\begin{metanote}
  add proof terms and backchaining to sequents and code
\end{metanote}
\begin{figure}
\newcommand{\XXi}{{\color{blue}{\Xi}}}
\[
\infer{\XXi\vdash t = t}
      {\eqExpert{\XXi}}
\qquad
\infer{\XXi\vdash \true}
      {\trueExpert{\XXi}}
\qquad
\infer{\XXi\vdash G_1\vee G_2}
      {\XXi'\vdash G_i\qquad \orExpert{\XXi}{\XXi'}{i}}
\qquad
\infer{\XXi\vdash \exists x. G}
      {\XXi'\vdash G[t/x]\qquad \someExpert{\XXi}{\XXi'}{t}}
\]
\vskip -6pt
\[
\infer{\XXi\vdash G_1\wedge G_2}
      {\XXi_1\vdash G_1\quad \XXi_2\vdash G_2\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}}
\qquad
\infer{\XXi\vdash A}
      {\XXi'\vdash G \quad (A~\hbox{\tt :-}~G)\in\instan\Pscr
                     \quad \unfoldExpert{\XXi}{\XXi'}}
\]
\caption{A proof system augmented with proof certificates and
  additional ``expert'' premises.}
\label{fig:augmented}
TODO
%\lstinputlisting[linerange={sigs-end}]{code/kernel.sig}
%\lstinputlisting[linerange={check-end}]{code/kernel.mod}
\caption{A simple proof checking kernel.}
\label{fig:kernel}
\end{figure}

Figure~\ref{fig:augmented} contains a simple proof system for Horn
clause provability in which each inference rule is augmented with an
additional premise involving an \emph{expert predicate}, a
certificate $\Xi$, and possibly continuations of certificates ($\Xi'$,
$\Xi_1$, $\Xi_2$) with extracted information from certificates (in the
case of $\vee$ and $\exists$).
%
%
% 
Although this proof system is a focused proof system, the
richness of focusing is not apparent in this simplified 
setting: thus, we drop the adjective ``focused'' from this point forward.

Figure~\ref{fig:kernel} contains the \lP implementation of the
inference rules in Figure~\ref{fig:augmented}: here the infix
turnstile $\vdash$ symbol is replaced by the \lst<i{check} predicate.
%
Notice that it is easy to show that no matter how the expert predicates
are defined, if the goal \lsti{check Cert B} is provable in \lP then
\lsti{B} must be a sound consequence of the program clauses stored in
\dots
% the \lsti{prog} predicate (which provides a natural implementation of
% the premise $(A~\hbox{\tt :-}~G)\in\instan{\Pscr}$.
% AM: not there
% in Figure~\ref{fig:kernel}).

@@

\begin{metanote}
  Rethink and update (with the right connectives) these FPC
\end{metanote}
\begin{figure}
\lstinputlisting[linerange={resources-end}]{code/fpcs.sig}
\lstinputlisting[linerange={resources-end}]{code/fpcs.mod}
\caption{Two FPCs that describe proofs that are limited in either
  height or in size.}
\label{fig:resources}
\end{figure}

@@

Various  additional examples and experiments using the pairing of FPCs
can be found in~\cite{Pair}. Using similar techniques, it is possible to define FPCs that target
specific types for special treatment: for example, when generating
integers, only (user-defined) small integers could be produced.
